The Heavy Photon Search Test Detector

### 2013-10-25 hps-testrun-nim.tex
Johns updated to abstract and intro. Some small fixes as well.

\begin{abstract}
Recent astrophysical data motivates searches for a massive, dark sector analog to the photon. The 
preferred parameter space for this "heavy photon" requires very high luminosity for direct production, 
favoring fixed target experiments, and results in long-lived decays of these new force carriers to pairs of 
light, charged fermions.  Identification of these states on large backgrounds requires precision tracking and 
vertexing at high rates in a difficult experimental environment. The Heavy Photon Search, or HPS, 
proposes a novel silicon tracking and vertexing detector, the HPS Silicon Vertex Tracker (SVT), to search 
for these states in a fixed target experiment at Thomas Jefferson National Accelerator Facility (JLab). 
The HPS SVT employs actively cooled silicon 
microstrip sensors with fast readout directly downstream of a target and inside of a dipole magnet to 
instrument a large acceptance with a relatively small detector. The readout is triggered by a fast, 
radiation-hard, compact, lead tungstate electromagnetic calorimeter.  As a first step; a simpler, small scale 
version of this detector has been assembled and operated on beam at JLab. A detector description and 
results on multiple Coulomb scattering models are outlined in this paper.
\end{abstract}

@@

There are two necessary parameters that determine the properties of this ``heavy photon," designated the 
\Aprime.  The first is the mixing of this state with the Standard Model photon, $\epsilon$, which results in 
a coupling of the \Aprime to Standard Model fermions with a strength of $\epsilon e$.  In simple models 
with kinetic mixing, $\epsilon$ is favored to be in the range of $10^{-2} -10^{-5}$.  The second is the mass 
of the \Aprime.  Under the assumption that this mass is generated via a Higgs-like symmetry breaking, 
the mass would be roughly of the order $\sqrt{\epsilon} m_W \sim$~-GeV.  These ranges are further 
motivated by explanations of other dark matter anomalies involving non-minimal dark sectors~
\cite{TuckerSmith:2001hy} and the fact that no similar excess in proton-antiproton pairs is observed, which 
would be simply explained if $ m_{\textrm{A}^\prime} < 2m_p$~\cite{Adriani:2010rc}.

Searches for these heavy photons with high-intensity, fixed target experiments offer the best sensitivity in 
this region of the parameter space~\cite{Bjorken:2009mm}.  In these experiments, an \Aprime is 
radiated off the incident electron, much like a bremsstrahlung photon, which subsequently decays to a pair 
of Standard Model fermions.  However, achieving sensitivity to this parameter space presents a number of 
challenges.  At larger couplings, large data samples are required to identify a small invariant mass peak on 
significant background.  At smaller couplings, the  \Aprime becomes long-lived, which enables 
sensitivity from extremely high-luminosity beam-dump experiments that screen out the Standard Model 
backgrounds.  However, in the intermediate region, lifetimes are too short to screen out backgrounds with a 
shield, but signal cross sections are so small that collecting large enough event samples to distinguish the 
signal requires unacceptable data rates or running times.  Gaining sensitivity to this region requires a fast 
detector capable of precision mass reconstruction of \Aprime final states together with the best possible 
rejection of prompt vertexes to eliminate backgrounds. The HPS experiment aims to achieve these goals 
by placing a compact silicon and vertexing detector ($\approx 1$ m long) immediately downstream ($
\approx 10$ cm) of a thin target ($\approx0.25$ X$_0$ tungsten) and inside of a dipole magnet ($\approx 
1$ T)~\cite{HPS_proposal_2010}.

The HPS Test run was proposed to DOE early in 2011 as the first stage of the HPS experiment. Its
purposes included demonstrating that the apparatus and data acquisition systems are feasible and
that the trigger rates and occupancies encountered in electron-beam running are as simulated.  It also
provided valuable experience to the HPS Collaboration in all aspects of designing, building, installing,
and running the experiment at JLab. Furthermore, in the case that HPS Test detector met all of the 
performance goals and was given dedicated running time with an electron beam, the HPS Test Run 
could provide new sensitivity to heavy photons. The HPS Test apparatus was installed on April 19, 2012, 
and 
ran parasitically with the HDice experiment, using a photon beam, until May 18. The JLab run schedule 
precluded any dedicated electron beam running, but the HPS Test Run was allowed a short and valuable
dedicated run with the photon beam.

This paper reviews the HPS Test Run apparatus, a simplified version of that 
planned for the full HPS experiment, and demonstrates the feasibility of the detector
technologies proposed for silicon tracker, ECal, and data acquisition systems. It documents the
performance of the trigger, data acquisition, silicon tracker, and ECal
and shows that the performance assumed in calculating the physics reach of the experiment is realistic.
Of particular importance, data from dedicated photon beam running has been used to compare
the observed trigger rates with that expected in simulation. The trigger rate is almost entirely due to
photons which have converted to \ee upstream of HPS and is sensitive to the multiple
Coulomb scattering of electrons and positrons in the conversion target.  Since scattered primary beam is 
the dominant 
source of occupancy in running HPS in an electron beam, good agreement between data and simulation 
confirm the background simulation used to benchmark the physics reach of the HPS experiment.

The very small signal rate compared to the large background, the need to fully reconstruct the final states of 
the electron and positron and doing it in a dense radiation environment millimeters from a high-current beam place stringent requirements on the detector, which should have,
\begin{itemize}
\item large and uniform acceptance in the forward region to catch boosted decay products,
\item excellent reconstruction efficiency for electron and positrons  with energy down to $XXXX$~MeV,
\item good momentum resolution to separate a promptly decaying \Aprime signal from the irreducible background,
\item excellent vertex resolution along the beam line direction to search for displaced \Aprime decays,
\item low-noise electronics with event timing resolution down to 2~ns, 
\item a flexible, redundant and efficient trigger selecting electron and positron pairs at rates up to 50kHz,
\item an online computing framework and data acquisition capable of handling event rates of 50~kHz and 
bandwidths of 100~MB/s to permanent storage,
\item detector components that can survive and efficiently operate in a high radiation environment with 
local doses exceeding $XXXX$~MRad
\end{itemize}

@@

A realistic {\sc Geant4} simulation of the detector based on the {\sc SLIC}~\ref{slic} simulation 
package~\cite{Graf:2011zzc}
including all backgrounds has been used to optimize the layout for acceptance, tracking efficiency, mass 
resolution and prompt vertex rejection. The layout of the SVT is summarized in Tab.~\ref{tab:trk} and rendered in Fig.~\ref{fig:tracker_model}.

@@

\subsubsection{Calibration}
\label{sec:ecal_calibration}
The noise and pedestal of the readout chain are calibrated by running the ECal readout in a mode where the 
preamplifier output is sampled every 4~ns in a time window of 100 samples: by looking at a part of the 
window before the hit, we calibrate the readout channel. The The readout threshold

>>>>

\begin{abstract}
Heavy photons, aka "hidden sector" or "dark" photons, are massive cousins to the SM photon, 
which couple to an analogue of electric charge. They arise naturally in theories with additional
U(1) gauge groups, and have been proposed as a possible explanation for the muon g-2 anomaly,
astrophysical puzzles, and possible sightings of light dark matter in direct detection experiments.
Masses in the range 10 to 1000 MeV are suggested. Heavy photons couple weakly to electric charge, 
so can be produced via electroproduction and can decay into elecron-positron pairs, albeit at rates 
strongly suppressed compared to standard QED trident processes. Consequently, heavy photons must be identified
above copious backgrounds, requiring high luminosity/high rate experiments, good mass resolution
(to identify invariant mass bumps), and good vertexing (to see the long lifetimes implied by
weak couplings). The Heavy Photon Search (HPS) will use  a silicon microstrip vertex tracker
triggered by a high speed PbWO4 electromagnetic calorimeter to search 
for these states in a fixed target experiment at Thomas Jefferson National Accelerator Facility (JLab). 
The first stage of HPS was the HPS Test Run experiment, designed to demonstrate the technical feasibility of the detector and data
acquisition systems and to confirm that critical backgrounds were correctly modeled.
This paper describes the HPS Test Run detector, reviews its performance, and presents results of
the normalized trigger rate measurements that confirm the accuracy of the background simulation.
\end{abstract}

@@

Heavy photons, aka known as "hidden sector" or "dark" photons, are particles with mass 10-1000 MeV which 
couple
weakly to the SM photon, so can be radiated by electrons, and can decay into e+e- pairs, albeit at rates far
below those of QED trident processes. They have been suggested by numerous Beyond Standard Model 
theories and
as an explanation of the current muonic g-2 anomaly, and may couple directly
to hidden sector particles that may constitute all or some of the dark matter. Current phenomenology highlights 
the 
20-1000 MeV mass range, and suggests that heavy photons couple to electric charge with strength 
$\epsilon$e, where 
$\epsilon$ is in the range of $10^{-3} -10^{-5}$. This range of parameters makes heavy photon searches viable 
at electron-positron colliders and in moderate energy fixed target electroproduction, and results from both
approaches have already been reported. Fixed target experiments offer the highest sensitivity 
for heavy photons with mass $<1000$~MeV~\cite{Bjorken:2009mm}, but require large data sets and good 
mass resolution to identify a small mass peak above the copious QED background. At small couplings, heavy 
photons become long-lived, so detection of a secondary decay vertex can reject the prompt QED background 
and boost experimental sensitivity.
The HPS experiment uses both invariant mass and vertex signatures by placing a compact silicon tracking and 
vertexing detector ($\approx 1$~m long) immediately downstream ($\approx 10$~cm) of a thin target 
($\approx0.25~X_0$ tungsten) and inside of a dipole magnet ($\approx 1$~T)~\cite{HPS_proposal_2010}. 
The experiment runs at high rates, exploiting the 100\% duty cycle of the 
CEBAF accelerator to accumulate the need statistics. The trigger is provided by a fast electromagnetic 
calorimeter just downstream of the magnet.

The HPS Test run, a simplified version of the full HPS experiment, was proposed and approved at JLab 
as the first stage of the HPS experiment.  Its purposes included demonstrating that the apparatus and data 
acquisition systems are technically feasible and
that the trigger rates and occupancies encountered in electron-beam running are as simulated. Additionally,
given dedicated running time with electron beams, the HPS Test Run could provide new sensitivity to heavy 
photons. The HPS Test apparatus was installed on April 19, 2012, and 
ran parasitically with the HDice experiment, using its photon beam, until May 18. The JLab run schedule 
precluded any dedicated electron beam running, but the HPS Test Run was allowed a short and valuable
dedicated run with the photon beam.

This paper reviews the HPS Test Run apparatus. It documents the successful performance of the trigger, data 
acquisition, silicon tracker, and ECal and
shows that the performance assumed in calculating the physics reach of the experiment is realistic.
Of particular importance, data from the dedicated photon beam running has been used to compare
the measured trigger rates with those expected in simulation. The Test Run trigger rate is almost entirely due to
photons which have converted to \ee pairs in a thin target upstream of HPS and it is sensitive to the particulars 
of their multiple Coulomb scattering in the conversion target.  Since, during electron beam running, the tails of 
the multiply Coulomb scattered primary beam are the dominant source of occupancy in the tracker and the 
trigger rate in the Ecal, they must be understood quantitatively. The trigger rates in the Test Run are also 
sensitive to the tails of the multiple Coulomb scattering distribution. Good agreement between the measured 
and simulated rates confirms our understanding of multiple Coulomb scattering and 
the reliability of the background simulation used to benchmark the physics reach of the HPS experiment.

The very small signal rate compared to the large background and the need to reconstruct the final states of 
the electron and positron in a dense radiation environment millimeters from a high-current beam 
place stringent requirements on the detector, which should have,
\begin{itemize}
\item large and uniform acceptance in the forward region to catch boosted decay products,
\item a flexible, redundant and efficient trigger selecting electron and positron pairs at rates up to 50kHz,
\item excellent track reconstruction efficiency for electrons and positrons,
\item good angular and momentum resolution to reconstruct invariant mass precisely,
\item excellent vertex resolution to search for displaced heavy photon decays,
\item high rate electronics with event timing resolution down to 2~ns,
\item data acquisition system capable of handling data rates of 100~MB/s to permanent storage,
\item detector components that can survive and efficiently operate in a high radiation environment with 
local doses exceeding $100$~MRad
\end{itemize}

@@

A realistic {\sc Geant4} simulation of the detector based on the {\sc SLIC} simulation 
package~\cite{Graf:2011zzc}
including all backgrounds has been used to optimize the layout for acceptance, tracking efficiency, mass 
resolution and prompt vertex rejection. The layout of the SVT is summarized in Tab.~\ref{tab:trk} and rendered in Fig.~\ref{fig:tracker_model}.

@@

%\subsubsection{Calibration}
%\label{sec:ecal_calibration}
The noise and pedestal of the readout chain are calibrated by running the ECal readout in a mode where the 
preamplifier output is sampled every 4~ns in a time window of 100 samples: by looking at a part of the 
window before the hit, we calibrate the readout channel. The The readout threshold
